Basics of Hardware Cache Optimization

### 2017-06-13 tex/abstract-.tex
Tweak

We have seen that
\alts{
  {the hidden constant \alts{separating, distinguishing} the time complexities of two
  reasonable algorithms under asymptotic analysis can get quite big in the presence of a
  memory hierarchy.},
  {analyzing \alts{an algorithm's,} \alts{time, asymptotic, algorithmic} complexity,
  % This is not true: it doesn't matter for asymptotic complexity if there are different
  % types of memory.
  \x{under the usual \gls{ram} model, where a single uniform memory is \alts{presumed,
  assumed, supposed, premised},}
  can be quite inaccurate in the presence of a memory hierarchy.},
}
\alts{
  To \alts{escape, avert, prevent} having to rely \alts{only, entirely, exclusively} on
  empirical results, To overcome this problem,
},
abstract machine models taking the non-uniform \alts{memories of, nature of memory in}
real-world computers into account \alts{can be used, are used, have been proposed}.
% By whom?
One of these is the \gls{emm}\alts{
  ,
  ~\cite[5]{afmh},
  {, used \alts{in~\cite{afmh}, by \textcite{afmh}}.},
}.

\subsection{External Memory Model}

% https://en.wikipedia.org/wiki/Random-access_machine

% TODO: go into more detail about the RAM model?
\alts[2]{
  {\alts{Establishing the basis of the, The} \gls{emm}, the widely used \gls{ram} model
  assumes \say{a \say{sufficiently} large uniform memory}~\cite[5]{afmh} with a constant
  access time.
  The},
  {The \gls{emm} is an extension of the \gls{ram} model.  While the latter assumes \say{a
  \say{sufficiently} large uniform memory}~\cite[5]{afmh} with a constant access time,
  the},
}
\gls{emm} \alts{divides, \alts{expands upon, extends} this by dividing} the memory into
\emph{internal} and \emph{external}.
\alts{The, Only the} internal memory is accessed directly, but its size is limited to
\(M\) \alts{items, words}.
The external memory is unbounded, but \alts{
  can only be accessed indirectly by loading data,
  data \x{is not accessed directly and} has to be loaded,
}
into internal memory \say{using \glslink{io}{I/Os} that move \(B\) contiguous
\alts{[items], words}}~\cite[5]{afmh}.

% > Although the word "I/O" suggests that external memory should be identified with disk
% > memory, we are free to choose any two levels of the memory hierarchy for internal and
% > external memory in the model.
%        -- Algorithms for Memory Hierarchies, page 6
The use of the term \emph{\acrshort{io}} here is somewhat non-standard.  While it suggests
\x{\alts{that, the}} external memory represents an \gls{hdd}, \gls{ssd} or similar, it is
not constrained which \alts{physical storages, members of an actual memory hierarchy} are
\alts[2]{represented by, associated \alts{with, to}} \x{the} internal and external memory.
% Which members of an actual memory hierarchy are \alts{represented by, associated to} the
% internal and external memory is \alts{not constrained, as desired}.
If we choose the set of all CPU caches and the main memory, \(B\) becomes the cache line
size and \(M\) may be in the order of a few \si{\mebi\byte}.

% > The relative miss penalty is much lower for caches or the TLB than for disks, so
% > constant factors are important, we find that asymptotic analysis is not enough to
% > determine the performance of algorithms.
%      -- Algorithms for Memory Hierarchies, page 181

% > In the EMM it is generally assumed that the cost of I/Os is much greater than the cost
% > of computation, hence the design of the algorithm is usually motivated by the need to
% > minimise the number of I/Os.
%      -- Algorithms for Memory Hierarchies, page 188

% \subsubsection{Caveats}

The number of \acrshortpl{io} an algorithm requires in the \gls{emm} can augment the
information provided by standard asymptotic complexity analysis%
\x{, if \acrshort{io} is pivotal to an algorithm's runtime},
but \alts[2]{can't substitute, is no substitute for} measurements.
For example, the lower bound of \acrshortpl{io} needed for computing the sum of some
input\x{, like in \cref{lst:ithare},}\ is \(\nicefrac{N}{B}\) in the \gls{emm}.  The
linked list in \cref{lst:ithare} probably takes almost \(N\) \acrshortpl{io}, since
consecutive \alts{nodes, elements} are unlikely to fall into the same cache line.  Thus,
the maximum predicted performance difference between \mintinline{cpp}{std::vector} and
\mintinline{cpp}{std::list} is \(B\), the number of items a cache line can hold, which is
\si{16} in this case.%
\footnote{%
  % TODO: elaborate on the size of `int`?
  A cache line has \alts{\si{64}, \SI{64}{\byte}} and an \mintinline{text}{int} \si{4}
  bytes on the machine used.
}
Recalling that the actual performance difference was \num[round-mode=places,
round-precision=0]{\speedup}, this is pretty inaccurate, but more informative than saying
both algorithms' time complexities are \x{in} \(\Theta(N)\).

% This analysis only makes sense when \acrshort{io} is pivotal to an algorithm's
% efficiency.

% The objective is to \alts{
%   {design algorithms requiring a \alts{minimum number, minimum, minimal number} of
%   \acrshortpl{io}.},
%   {minimize the number of \alts{\acrshortpl{io}, I/Os} \alts{an algorithm needs,
%   needed}.},
% }

\x{
% Applying this model to \cref{lst:ithare} suggests
The \gls{emm} suggest that we can at best sum containers like in \cref{lst:ithare} using
\(\nicefrac{N}{B}\) \acrshortpl{io}, where \(N\) is the number of items to sum and \(B\)
the amount of them that fit in one block.
% Since there's 5000 containers of size 5000 and my cache line
% size is \SI{64}{B} and fits 8 \mintinline{cpp}{int} values:
% \begin{equation*}
%   N/B = 5000^2/8 = 3125000
% \end{equation*}
We can assume that the version with the list uses almost \(N\) \acrshortpl{io} because of
the random memory access.  Since a cache line fits 16 \mintinline{text}{int} values (\(B =
16\)) it
predicts a performance difference of 16.  This is not very accurate.
Analysis of \acrshortpl{io} needed by an algorithms \x{or data structure} may help inform
the decision of whether to consider it for empirical comparisons, though.
% that can be used prior to empirical comparisons.

% Apparently, the model is not practical for comparing these two data structures.
}

% Now for something completely different.
\x{
The \gls{emm} also suggests another approach to improve \cref{lst:ithare}, if we don't
want to \alts{trade, forgo, give up} the \x{asymptotically} constant time complexity of
insertions and deletions anywhere in the container: an \emph{unrolled list}.
}

% Notes and potential problems:
% * The concept of I/Os that move B contiguous words directly translates to cache lines.
% * There's nothing equivalent to prefetching in the EMM.
% * > The main problem with hardware caches is that they use a fixed simplistic strategy
%   > for deciding which blocks are kept whereas the external memory model gives the
%   > programmer full control over the content of internal memory.
%        -- Algorithms for Memory Hierarchies, page 6
% * We lost associativity too.
% * And we don't model multiple levels of CPU caches, of course.

% > In the EMM it is generally assumed that the cost of I/Os is much greater than the cost
% > of computation, hence the design of the algorithm is usually motivated by the need to
% > minimise the number if I/Os.  Howeever in the CMM and the IMM the relative miss
% > penalties are far smaller and we have to consider computation costs.
%      -- Algorithms for Memory Hierarchies, Section 8.6.4, page 188

% > In general it is best to hardcode cache line sizes at compile time by using the
% > `getconf` utility as in:
% >    gcc -DCLS=$(getconf LEVEL1_DCACHE_LINESIZE) ...
% > If the binaries are supposed to be generic, the largest cache line size should be
% > used.
%      -- Drepper, p. 50

% \subsubsection{Notes}
\subsubsection{Limitations and Other Models}

% Replacement policies.  Associativity.  More than two levels of memory.

% While \alts{
While the concept of \acrshortpl{io} directly models \alts{
  most\footnote{%
    TODO: not associativity though.
  },
  the
}
\alts{effects, existence} of cache lines,
% are directly modeled by the concept of \acrshortpl{io},
nothing in the \gls{emm} accounts for \x{those of} prefetching.

\input{tex/com.tex}

>>>>

We have seen that
\alts{
  {the hidden constant \alts{separating, distinguishing} the time complexities of two
  reasonable algorithms under asymptotic analysis can get quite big in the presence of a
  memory hierarchy.},
  {analyzing \alts{an algorithm's,} \alts{time, asymptotic, algorithmic} complexity,
  % This is not true: it doesn't matter for asymptotic complexity if there are different
  % types of memory.
  \x{under the usual \gls{ram} model, where a single uniform memory is \alts{presumed,
  assumed, supposed, premised},}
  can be quite inaccurate in the presence of a memory hierarchy.},
}
\alts{
  To \alts{escape, avert, prevent} having to rely \alts{only, entirely, exclusively} on
  empirical results, To overcome this problem,
},
abstract machine models taking the non-uniform \alts{memories of, nature of memory in}
real-world computers into account \alts{can be used, are used, have been proposed}.
% By whom?
One of these is the \gls{emm}\x{\alts{
  ~\cite[5]{afmh},
  {, used \alts{in~\cite{afmh}, by \textcite{afmh}}.},
}}.

\input{tex/emm}
\input{tex/com}

### 2017-06-13 tex/emm.tex
Tweak

\section{Abstract?}
% \section{Analysis of Algorithms}

% > However the efficiencies of any two "reasonable" implementations of a given algorithm
% > are related by a constant multiplicative factor called a *hidden constant*.
%      -- https://en.wikipedia.org/wiki/Analysis_of_algorithms

% Mathematical analysis of algorithms to obtain asymptotic upper bounds is nice.  Ignoring
% the non-uniform nature of memory hierarchies can yield pretty bad results, though.  What
% can be done about this?

We have seen that
\alts{
  {the hidden constant \alts{separating, distinguishing} the time complexities of two
  reasonable algorithms under asymptotic analysis can get quite big in the presence of a
  memory hierarchy.},
  {analyzing \alts{an algorithm's,} \alts{time, asymptotic, algorithmic} complexity,
  % This is not true: it doesn't matter for asymptotic complexity if there are different
  % types of memory.
  \x{under the usual \gls{ram} model, where a single uniform memory is \alts{presumed,
  assumed, supposed, premised},}
  can be quite inaccurate in the presence of a memory hierarchy.},
}
\alts{
  To \alts{escape, avert, prevent} having to rely \alts{only, entirely, exclusively} on
  empirical results, To overcome this problem,
},
abstract machine models taking the non-uniform \alts{memories of, nature of memory in}
real-world computers into account \alts{can be used, are used, have been proposed}.
% By whom?
One of these is the \gls{emm}\alts{
  ,
  ~\cite[5]{afmh},
  {, used \alts{in~\cite{afmh}, by \textcite{afmh}}.},
}.

@@

% > Although the word "I/O" suggests that external memory should be identified with disk
% > memory, we are free to choose any two levels of the memory hierarchy for internal and
% > external memory in the model.
%        -- Algorithms for Memory Hierarchies, page 6
The use of the term \emph{\acrshort{io}} here is somewhat non-standard.  While it suggests
\x{\alts{that, the}} external memory represents an \gls{hdd}, \gls{ssd} or similar, it is
not constrained which \alts{physical storages, members of an actual memory hierarchy} are
\alts[2]{represented by, associated \alts{with, to}} \x{the} internal and external memory.
% Which members of an actual memory hierarchy are \alts{represented by, associated to} the
% internal and external memory is \alts{not constrained, as desired}.
If we choose the set of all CPU caches and the main memory, \(B\) becomes the cache line
size and \(M\) may be in the order of a few \si{\mebi\byte}.

@@

The number of \acrshortpl{io} an algorithm requires in the \gls{emm} can augment the
information provided by standard asymptotic complexity analysis%
\x{, if \acrshort{io} is pivotal to an algorithm's runtime},
but \alts[2]{can't substitute, is no substitute for} measurements.
For example, the lower bound of \acrshortpl{io} needed for computing the sum of some
input\x{, like in \cref{lst:ithare},}\ is \(\nicefrac{N}{B}\) in the \gls{emm}.  The
linked list in \cref{lst:ithare} probably takes almost \(N\) \acrshortpl{io}, since
consecutive \alts{nodes, elements} are unlikely to fall into the same cache line.  Thus,
the maximum predicted performance difference between \mintinline{cpp}{std::vector} and
\mintinline{cpp}{std::list} is \(B\), the number of items a cache line can hold, which is
\si{16} in this case.%
\footnote{%
  % TODO: elaborate on the size of `int`?
  A cache line has \alts{\si{64}, \SI{64}{\byte}} and an \mintinline{text}{int} \si{4}
  bytes on the machine used.
}
Recalling that the actual performance difference was \num[round-mode=places,
round-precision=0]{\speedup}, this is pretty inaccurate, but more informative than saying
both algorithms' time complexities are \x{in} \(\Theta(N)\).

@@

\input{tex/com.tex}

% [1]: https://en.wikipedia.org/wiki/Analysis_of_algorithms

>>>>

@@

% > Although the word "I/O" suggests that external memory should be identified with disk
% > memory, we are free to choose any two levels of the memory hierarchy for internal and
% > external memory in the model.
%        -- Algorithms for Memory Hierarchies, page 6
The use of the term \emph{\acrshort{io}} here is somewhat non-standard.  While it suggests
\x{\alts{that, the}} external memory represents an \alts{
  {\gls{hdd} or \gls{ssd}},
  {\gls{hdd}, \gls{ssd} or similar},
}, it is
not constrained which \alts{physical storages, members of an actual memory hierarchy} are
\alts[2]{represented by, associated \alts{with, to}} \x{the} internal and external memory.
% Which members of an actual memory hierarchy are \alts{represented by, associated to} the
% internal and external memory is \alts{not constrained, as desired}.
If we choose the set of all CPU caches and the main memory, \(B\) becomes the cache line
size and \(M\) may be in the order of a few \si{\mebi\byte}.

@@

The number of \acrshortpl{io} an algorithm requires in the \gls{emm} can augment the
information provided by standard asymptotic complexity analysis%
\x{, if \acrshort{io} is pivotal to an algorithm's runtime},
but \alts[2]{can't substitute, is no substitute for} measurements.
For example, the lower bound of \acrshortpl{io} needed for computing the sum of some
input, like in \cref{lst:ithare}, is \(\nicefrac{N}{B}\) in the \gls{emm}.  The
linked list in that example probably takes almost \(N\) \acrshortpl{io}, since
consecutive \alts{nodes, elements} are unlikely to fall into the same cache line.  Thus,
the predicted performance difference between \mintinline{cpp}{std::vector} and
\mintinline{cpp}{std::list} is at most \(B\), the number of items a cache line can hold,
which is \si{16} in this case.%
\footnote{%
  % TODO: elaborate on the size of `int`?
  A cache line has \alts{\si{64}, \SI{64}{\byte}} and an \mintinline{text}{int} \si{4}
  bytes on the machine used.
}
Recalling that the actual performance difference was \num[round-mode=places,
round-precision=0]{\speedup}, this is pretty inaccurate, but more informative than saying
both algorithms' time complexities are \x{in} \(\Theta(N)\).

@@
